### 2021-02-10 docs/manuscript/PLOS-MSc-Thesis.tex
Improve introduction, figures and references

\subsection*{Conclusions}

@@

Cells interact continuously with their environment and they must detect signals
from it. In many cases Gene Regulatory Networks (GRNs) interpret these signals
through gene interaction and produce a particular response. This information
processing function is of particular importance in developmental processes such
as those in which a group of cells differentiate in response to a signalling
molecule. Such molecules were referred to as morphogens by Turing in 1952
\cite{Turing1952} and posterior theoretical studies on morphogens led to the
french flag model \cite{Wolpert1969}, in which a gradient of concentration of
the morphogen induces differentiation in a one-dimensional field of cells.
Although the Bicoid protein (bcd), the first example of a molecule that acted
as a morphogen, was only found in the 80's in the developing embryos of
\textit{Drosophila melanogaster}
\cite{driever_gradient_1988, driever_bicoid_1988}, now it is known that there
are many other examples such as the Decapentaplegic (Dpp) protein in
\textit{Drosophila} wing imaginal discs \cite{affolter_decapentaplegic_2007},
as well as Hedgehog and Wnt.
However, information processing by GRNs has proven to be a complex process, and
a major goal of developmental biology is to understand mechanistically how
positional information given by morphogens is translated in spatial
differentiation.\\

Our understanding of how GRNs process information has increased thanks to the
concept and theory of network motifs, which are defined as patterns of
interconnections occurring in networks at numbers that are significantly higher
than those in randomized networks \cite{Milo2002}. This novel approach, along
with mathematical modeling and computational methods have been shown to be a
powerful tool to study GRNs and the predictions of studies using these methods
have been validated by experimental studies
\cite{Kalir2005,Kalir2004,Mangan2003a,ODonnell2005}, and some of them
have even used synthetic GRNs to produce artificial cell differentiation~\cite{Basu2005}
and circadian gene expression~\cite{atkinson_development_2003}. Moreover,
the design and implementation of GRNs through synthetic biology is emerging as
a promising tool to study biological phenomena as pattern formation
\cite{santos-moreno_using_2019}.

@@

The set of interaction values was represented as adjacency matrices $W \in
\mathbb{R}^{3\times4}$ where $i$ represents a regulated node of the GRN and $j$
represents a regulator node of the GRN Fig \ref{fig:adjmat} with
$w_{\textit{ij}}~\in [-10,\ 10]$. Values in these matrices can be negative, zero
or positive and represent represion, no interaction, or activation respectively.
The magnitude of the value is proportional to the strength of the interaction.\\

\begin{figure}[!h]
%  \centering
 \includegraphics[width=\textwidth]{figures/metodos/matriz-de-adyacencia}
    \caption{\bf Modelling of GRNs.}
    A: GRN represented as a directed graph with activation (blue arrows) and
    inhibition (red arrows) interactions. Black circles
    tagged with a \textbf{M} represent the morphogen; \textbf{A}, \textbf{B} and
    \textbf{C} represent genes in the network.
    B: The same GRN represented as an adjacency matrix in which positive values
    represent activation, negative ones represent inhibition and
    zeros represent no interaction.
 \label{fig:adjmat}
\end{figure}

In our model the morphogen could interact with any of the three genes on the
GRN, but could not be affected by them. The morphogenetic field was defined by
an unidimensional array of 30 isogenic cells exposed to the morphogen
concentration gradient (Fig~\ref{fig:morfgrad}). The initial concentration of
each gen product in all cells was set to 0.1 in all simulations.\\

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/metodos/morphogen-gradient}
 \caption{\bf Morphogenetic field.}
 In our model the field was composed of a linear array of 30 isogenic cells and
 a morphogen gradient concentration described by an exponential decay function.
 Horizontal arrows between cells represent diffusion of gene products between
 adjacent cells.
 \label{fig:morfgrad}
\end{figure}

@@

The optimal pattern of gene expression defined by us, consisted in cells at the
sides of the field ($n<11$ and $n>20$) displaying expression levels lower than
10\% of the maximal level observed along the field of cells for the output gene,
and cells at the middle of the field ($n ∈ [11,20]$) displaying expression
levels greater than 90\% of the maximal level observed along the entire field
for the output gene (Fig~\ref{fig:exp-pattern}).

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/metodos/optimal-pattern}
 \caption{\bf Optimal gene expression pattern.}
 \label{fig:exp-pattern}
\end{figure}

@@

\begin{figure}[!h]
 \center
 \includegraphics[width=\textwidth]{figures/results/distribution_topologies_en}
 \caption{\bf Distribution of abundances of network topologies.}
 \label{fig:distopol}
\end{figure}

The most abundant network topologies are shown in Fig~\ref{fig:topol}.
Interestingly, among the 15 most abundant network topologies, eleven of them
presented the Incoherent type 3 Feed-Forward Loop (I3-FFL) network motif. In
addition, we found that topology number 9 could produce the striped phenotype
with only two nodes.

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/most-abu-topol}
 \caption{\bf Most frequent network topologies.}
 Most frequent network topologies contain the I3-FFL network motif. Topologies
 are numbered in order of decreasing abundance.
 The number of GRNs belonging to each network topology is presented below
 topology number.
 Network topology number 9 produces the striped phenotype using a two-node GRN.
 \label{fig:topol}
\end{figure}

@@

The degree distribution of this network seems to follow a Poisson distribution
(Pearson’s $\chi^2 = 12.8902$, $\text{p-value} = 0.115684$) characteristic of a
Erdös-Rényi network \cite{Erdos1959}. This kind of network presents a great
number of nodes with low degree and a few nodes with high degree
(Fig~\ref{fig:deg-dist}). Nodes corresponding to topologies number one and three
showed the highest degree ($deg(1) = deg (3) = 14$). Curiously, nodes with
higher degree resulted to be also those network topologies with higher abundance
(Fig~\ref{fig:ab-deg}). This relationship between abundance and node degree
was corroborated by the Spearman's rank correlation coefficient
($r_S = −0.353752$, $p = 6.2933\times10^{-23} $).

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/neutral-network-degree-distribution}
 \caption{\bf Histogram of vertex degree distribution in the neutral network.}
 \label{fig:deg-dist}
\end{figure}

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/abundance-vs-node-degree}
 \caption{\bf Scatter plot of Topology abundance vs. Node degree in the neutral
 network.}
 \label{fig:ab-deg}
\end{figure}

@@

Besides topological robustness, we wanted to know if network topologies were
robust to perturbations in the values of the parameters. After performing the
perturbations to a GRN from each topology, we plotted the abundance of the
topology vs. its robustness. We did not find a significant correlation between
these two variables ($r_S = −0.0389724$, $\text{p-value} = 0.298475$)
(Fig~\ref{fig:ab-rob}). However, when we grouped those topologies containing
the same subgraph and calculated their mean robustness, we found that between
the most robust subgraphs is the ``Bistable'' motif reported by Cotterell \&
Sharpe \cite{Cotterell2010}, the I3-FFL and the ``Overlapping domains'' motif
(which is a variation of the I3-FFL) also reported by Cotterell \& Sharpe (see
Supplementary Material).

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/abundance-vs-robustness}
 \caption{\bf Scatter plot of Abundance vs. robustness.}
 The robustness of each topology was calculated as the number of perturbations
 in which the fitness was greater or equal to 0.95 with respect to the total of
 perturbations performed.
 \label{fig:ab-rob}
\end{figure}

@@

The average complexity index of network topologies was 39.2435 ($\pm 11.6768$)
and was found to be negatively correlated with the abundance ($r_S = -0.316653$,
$\text{p-value} = 2.25161\times10^{-18}$) (Fig~\ref{fig:ab-comp}).

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/abundance-vs-complexity-index-log}
 \caption{\bf Scatter plot of Abundance vs. Complexity index.}
 Each point in the plot represents one of the 714 network topologies.
 \label{fig:ab-comp}
\end{figure}

Hierarchical clustering of spatiotemporal expression profiles showed that GRNs
with topologies that seemed to be very different have indeed very similar
expression dynamics. We expected to find GRNs with the same topology to group
together but instead we found GRNs with different topologies grouped as close
neighbors (Fig~\ref{fig:exp-tree}). More interestingly, some GRNs displaying
very abundant network topologies were grouped together with GRNs displaying
very infrequent network topologies.

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/spatio-temporal-profile-clustering-i3-ffl}
 \caption{\bf Clustering of spatiotemporal expression profiles obtained by
 neighbor joining showing the presence of I3-FFL network motif.}
 \label{fig:exp-tree}
\end{figure}

Clustering of topologies revealed that there are six main groups of topologies
in our set of 2061 GRNs (Fig~\ref{fig:subg-tree}).

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/subgraph-profile-clustering-i3-ffl}
 \caption{\bf Clustering of network topologies.}
 Blue circles correspond to topologies that contain the I3-FFL network motif.
 \label{fig:subg-tree}
\end{figure}

@@

%PLOS does not support heading levels beyond the 3rd (no 4th level headings).
% \subsection*{\lorem\ and \ipsum\ nunc blandit a tortor}
% \subsubsection*{3rd level heading}
% Sed suscipit iaculis mi at feugiat.
%
% \begin{enumerate}
% 	\item{react}
% 	\item{diffuse free particles}
% 	\item{increment time by dt and go to 1}
% \end{enumerate}
%
%
% \subsection*{Sed ac quam id nisi malesuada congue}
%
% Praesent faucibus semper libero.
%
% \begin{itemize}
% 	\item First bulleted item.
% 	\item Second bulleted item.
% 	\item Third bulleted item.
% \end{itemize}

@@

That the most abundant topologies were found to be part of highly connected
clusters in the neutral network indicates that these topologies are both robust
to quantitative changes in the strength of interactions and diffusion and
degradation parameters as well as to qualitative changes in the topology
itself.

We found that the I3-FFL is one of the network motifs that appears more
frequently in these GRNs, as has also been reported by Cotterell \& Sharpe
\cite{Cotterell2010}, and has been found in transcriptional networks involved
in cell fate definition \cite{Li2019}. Moreover, our results indicate that this
is a robust motif that can be useful for construction of synthetic gene
regulatory networks that are exposed to stochastic fluctuations in the
environment and mutations.

@@

\section*{Conclusion}

%CO\textsubscript{2} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque. Quisque augue sem, tincidunt sit amet feugiat eget, ullamcorper sed velit.

@@

\paragraph*{S3 Fig.}
\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/supplementary-figures/GRN_660_topol_33_spatio-temp_exp_profile}
 \caption{\bf Most GRNs reach the steady state before 250 time steps.}
 GRN number 660 displaying the topology number 33 (A) and its spatiotemporal
 expression profile (B).
 \label{S3_Fig}
\end{figure}

% \paragraph*{S1 File.}
% \label{S1_File}
% {\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.
%
% \paragraph*{S1 Video.}
% \label{S1_Video}
% {\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.
%
% \paragraph*{S1 Appendix.}
% \label{S1_Appendix}
% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

>>>>

% \subsection*{Conclusions}

@@

Cells interact continuously with their environment and they must detect signals
from it. In many cases Gene Regulatory Networks (GRNs) interpret these signals
through gene interaction and produce a particular response. This information
processing function is of particular importance in developmental processes such
as those in which a group of cells differentiate in response to a signalling
molecule. Such molecules were referred to as morphogens by Turing in 1952
\cite{Turing1952} and posterior theoretical studies on morphogens led to the
french flag model \cite{Wolpert1969}, in which a gradient of concentration of
the morphogen induces differentiation in a one-dimensional field of cells.
Although the Bicoid protein (bcd), the first example of a molecule that acted
as a morphogen, was only found in the 80's in the developing embryos of
\textit{Drosophila melanogaster}
\cite{driever_gradient_1988, driever_bicoid_1988}, now it is known that there
are many other examples such as the Decapentaplegic (Dpp) protein in
\textit{Drosophila} wing imaginal discs \cite{affolter_decapentaplegic_2007},
as well as Sonic Hedgehog \cite{dessaud_dynamic_2010,cohen_theoretical_2014}
and Wnt \cite{raspopovic_digit_2014}. However, information processing by GRNs
has proven to be a complex process, and a major goal of developmental biology
is to understand mechanistically how positional information given by morphogens
is translated in spatial differentiation.\\

Our understanding of how GRNs process information has increased thanks to the
concept and theory of network motifs, which are defined as patterns of
interconnections occurring in networks at numbers that are significantly higher
than those in randomized networks~\cite{Milo2002}. This novel approach, along
with mathematical modeling and computational systems biology methods have been
shown to be a powerful tool to study GRNs~\cite{kitano_computational_2002}, and
the predictions of studies using these methods have been validated by
experimental studies~\cite{Kalir2005,Kalir2004,Mangan2003a,ODonnell2005}, and
some of them have even used synthetic GRNs to produce artificial cell
differentiation~\cite{Basu2005}, circadian gene
expression~\cite{atkinson_development_2003}, counting
devices~\cite{friedland_synthetic_2009} and systems that respond to light.
Moreover, the design and implementation of GRNs through synthetic biology is
emerging as a promising tool to study biological phenomena as pattern formation
\cite{santos-moreno_using_2019} and as a novel therapeutic tool with amazing
biomedical applications~\cite{karlsson_therapeutic_2012,
higashikuni_advancing_2017,abil_synthetic_2015,healy_genetic_2019,
kitada_programming_2018}.

@@

The set of interaction values was represented as adjacency matrices $W \in
\mathbb{R}^{3\times4}$ where $i$ represents a regulated node of the GRN and $j$
represents a regulator node of the GRN (Fig \ref{fig:model}A and
\ref{fig:model}B) with $w_{\textit{ij}}~\in [-10,\ 10]$. Values in these
matrices can be negative, zero or positive and represent represion, no
interaction, or activation respectively. The magnitude of the value is
proportional to the strength of the interaction.\\

\begin{figure}[!h]
%  \centering
 \includegraphics[width=\textwidth]{figures/metodos/fig_1}
    \caption{\bf Modelling of GRNs.}
    (A) GRN represented as a directed graph with activation (blue arrows) and
    inhibition (red arrows) interactions. Black circles
    tagged with a \textbf{M} represent the morphogen; \textbf{A}, \textbf{B} and
    \textbf{C} represent genes in the network.
    (B) The same GRN represented as an adjacency matrix in which positive values
    represent activation, negative ones represent inhibition and
    zeros represent no interaction.
    (C) Morphogenetic field. In our model the field was composed of a linear
    array of 30 isogenic cells and a morphogen gradient concentration described
    by an exponential decay function. Horizontal arrows between cells represent
    diffusion of gene products between adjacent cells. Optimal gene expression
    pattern is shown top right.
    (D) Spatiotemporal expression profile of a GRN.
 \label{fig:model}
\end{figure}

In our model the morphogen could interact with any of the three genes on the
GRN, but could not be affected by them. The morphogenetic field was defined by
an unidimensional array of 30 isogenic cells exposed to the morphogen
concentration gradient (Fig~\ref{fig:model}C). The initial concentration of
each gen product in all cells was set to 0.1 in all simulations.\\

@@

The optimal pattern of gene expression defined by us, consisted in cells at the
sides of the field ($n<11$ and $n>20$) displaying expression levels lower than
10\% of the maximal level observed along the field of cells for the output gene,
and cells at the middle of the field ($n ∈ [11,20]$) displaying expression
levels greater than 90\% of the maximal level observed along the entire field
for the output gene (Fig~\ref{fig:model}C).

@@

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/distribution_most_abundant_topologies}
 \caption{\bf Distribution of abundances of network topologies.}
 Most frequent network topologies shown in the grid contain the I3-FFL network
 motif. Topologies are numbered in order of decreasing abundance. The number of
 GRNs belonging to each network topology is presented below topology number.
 Note that network topology number 9 produces the striped phenotype using a
 two-node GRN.
 \label{fig:distopol}
\end{figure}

The most abundant network topologies are shown in Fig~\ref{fig:distopol}.
Interestingly, among the 15 most abundant network topologies, eleven of them
presented the Incoherent type 3 Feed-Forward Loop (I3-FFL) network motif. In
addition, we found that topology number 9 could produce the striped phenotype
with only two nodes.

@@

The degree distribution of this network seems to follow a Poisson distribution
(Pearson’s $\chi^2 = 12.8902$, $\text{p-value} = 0.115684$) characteristic of a
Erdös-Rényi network \cite{Erdos1959}. This kind of network presents a great
number of nodes with low degree and a few nodes with high degree
(Fig~\ref{fig:deg-dist}A). Nodes corresponding to topologies number one and
three showed the highest degree ($deg(1) = deg (3) = 14$). Curiously, nodes with
higher degree resulted to be also those network topologies with higher abundance
(Fig~\ref{fig:deg-dist}B). This relationship between abundance and node degree
was corroborated by the Spearman's rank correlation coefficient
($r_S = −0.353752$, $p = 6.2933\times10^{-23} $).

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/vertex_degree_dist-abundance_vs_node_degree}
 \caption{\bf Node degree in the neutral network.}
 (A) Histogram of vertex degree distribution in the neutral network.
 (B) Scatter plot of Topology abundance vs. Node degree in the neutral
 network.
 \label{fig:deg-dist}
\end{figure}

@@

Besides topological robustness, we wanted to know if network topologies were
robust to perturbations in the values of the parameters. After performing the
perturbations to a GRN from each topology, we plotted the abundance of the
topology vs. its robustness. We did not find a significant correlation between
these two variables ($r_S = −0.0389724$, $\text{p-value} = 0.298475$)
(Fig~\ref{fig:ab-rob}A). However, when we grouped those topologies containing
the same subgraph and calculated their mean robustness, we found that between
the most robust subgraphs is the ``Bistable'' motif reported by Cotterell \&
Sharpe \cite{Cotterell2010}, the I3-FFL and the ``Overlapping domains'' motif
(which is a variation of the I3-FFL) also reported by Cotterell \& Sharpe (see
Supplementary Material).

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/abundance-robustness-complexity-index}
 \caption{\bf Relationship between Abundance and robustness and complexity
 index.} (A) Abundance vs. Robustness. The robustness of each topology was
 calculated as the number of perturbations in which the fitness was greater or
 equal to 0.95 with respect to the total of perturbations performed.
 (B) Abundance vs. Complexity index. Each point in the plot represents one of
 the 714 network topologies.
 \label{fig:ab-rob}
\end{figure}

@@

The average complexity index of network topologies was 39.2435 ($\pm 11.6768$)
and was found to be negatively correlated with the abundance ($r_S = -0.316653$,
$\text{p-value} = 2.25161\times10^{-18}$) (Fig~\ref{fig:ab-rob}B).

Hierarchical clustering of spatiotemporal expression profiles showed that GRNs
with topologies that seemed to be very different have indeed very similar
expression dynamics. We expected to find GRNs with the same topology to group
together but instead we found GRNs with different topologies grouped as close
neighbors (Fig~\ref{fig:clustering}A). More interestingly, some GRNs displaying
very abundant network topologies were grouped together with GRNs displaying
very infrequent network topologies.

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/clustering_of_grns_and_topologies}
 \caption{\bf Clustering of expression profiles and topologies.}
 (A) Clustering of spatiotemporal expression profiles obtained by
 neighbor joining. (B) Clustering of network topologies. Blue circles
 correspond to topologies that contain the I3-FFL network motif.
 \label{fig:clustering}
\end{figure}

Clustering of topologies revealed that there are six main groups of topologies
in our set of 2061 GRNs (Fig~\ref{fig:clustering}B).

@@

@@

Robustness is an important feature of patterning mediated by morphogens
\cite{lo_robust_2015}. In our study, the most abundant topologies were found to
be part of highly connected clusters in the neutral network, and this fact
indicates that these topologies are both robust to quantitative changes in the
strength of interactions and diffusion and degradation parameters as well as to
qualitative changes in the topology itself. As robustness and evolvability are
not mutually exclusive properties \cite{wagner_robustness_2008}, it is likely
that these topologies are also evolvable.

Other studies have shown that Incoherent Feed Forward Loops are able to form
spatial stripes of gene expression \cite{ishihara_cross_2005}. In particular,
we found that the I3-FFL is one of the network motifs that appears more
frequently in these GRNs, as has also been reported by Cotterell \& Sharpe
\cite{Cotterell2010}, and has been found in transcriptional networks involved
in cell fate definition \cite{Li2019}. Moreover, our results indicate that this
is a robust motif that can be useful for construction of synthetic gene
regulatory networks that are exposed to stochastic fluctuations in the
environment and mutations.

@@

%\section*{Conclusion}

@@

\paragraph*{S3 Fig.}
\label{S3_Fig}
{\bf Most GRNs reach the steady state before 250 time steps.}
GRN number 660 displaying the topology number 33 (A) and its spatiotemporal
expression profile (B).

\paragraph*{S4 Fig.}
\label{S4_Fig}
{\bf Bridge topologies in the neutral network.}
%
% \paragraph*{S1 Video.}
% \label{S1_Video}
% {\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.
%
% \paragraph*{S1 Appendix.}
% \label{S1_Appendix}
% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.
